Predicting Federal Emergency Management Agency Funding for U.S. Disasters

### 2021-12-13 products/manuscript/Manuscript_Disasters.Rmd
Manuscript Updates - Part 1

# Introduction
On November 8, 1988, United States (US) President Ronald Reagan signed the Robert T. Stafford Disaster Relief and Emergency Act into law, which formalizes the disaster declaration process and specifies the response responsibilities for each tier of government [@moss_stafford_2009]. Under this law, when a disaster occurs, the affected area (local or state representatives) conducts a damage assessment to estimate the amount of support required from the federal government. That assessment goes through the state government to the Federal Emergency Management Agency (FEMA) to the Department of Homeland Security (DHS) to the US President, who ultimately decides whether the affected region should receive an emergency declaration, a major disaster declaration, or neither [@mccarthy_femas_2009]. FEMA then deploys resources from across the federal government to support the disaster region, the volume and type of which depend upon the declaration type. However, the federal government does not necessarily incur all of the cost of these resources. Under most circumstances, this pay structure more accurately reflects a reimbursement model; that is, the states or territories submit requests to the FEMA for reimbursement after the expenses have occurred. However, there is then a reconciliation process by which the total funds a state or territory receives is often different than the original request. The funding process was further refined as part of the Sandy Recovery Improvement Act of 2013 as well as The Disaster Recovery Reform Act of 2019 [@fema_guide_2021]. These laws require FEMA to publicly provide the information regarding federal government resources paid for by state and federal government funds for each disaster declaration. As a result, the OpenFEMA Data Sets website now exists and provides information summarizing disaster declarations and mission assignments. While this has improved the accessible information, little has been done to draw conclusions from these data for predicting the financial impacts of disasters.

@@

## Machine Learning Models
There are numerous machine learning models that could be applied to the research question, but this analysis focuses primarily on three different types of tree models. This type of model was chosen as the primary users of the prediction algorithms are non-statisticians, and tree models are among the easiest machine learning models to interpret and understand [@james_introduction_2021].

* Bagged Tree (BT): A slightly more sophisticated model, a BT aims to reduce the variance by re-sampling the existing data (otherwise known as bootstrapping) and builds a tree for each sample. The final model then averages all of the individually sampled trees. However, the reducing the variance in the model also reduces the ease of interpretation of the model [@dietterich_experimental_nodate].

* Random Forest (RF): A RF model follows similar principles to the BT, but instead of considering all possible predictors at each decision tree split, the split occurs based on the best predictors of a random sample of all predictors. This method allows for de-correlation in the trees by avoiding the default nature of decision trees that aim to include as many predictors as possible. Each individual tree is then averaged to find the final model. RF models are also more difficult to interpret than standard decision tree models [@breiman_random_2001].

@@

## Model Development
Three models were considered for the prediction analysis: decision tree, bagged tree, and random forest, and each were fit for both outcomes of interest (e.g. requested and obligated funds). The data were split such that 66.7% were used for training the models and the remaining 33.3% were used for testing the final chosen model, stratified by the incident type. Cross-validation was used for the decision tree and random forest models, using a five-fold, five-repeat resampling structure, also stratified by the incident type. The bagged tree model instead implemented a bootstrap resampling structure, repeated ten times, again stratified by the incident type. Incident type was chosen as the stratification variable to address the fact that the frequency of incident types varies across FEMA regions.

>>>>

# Introduction
On November 8, 1988, United States (US) President Ronald Reagan signed the Robert T. Stafford Disaster Relief and Emergency Act into law, which formalizes the disaster declaration process and specifies the response responsibilities for each tier of government [@moss_stafford_2009]. Under this law, when a disaster occurs, the affected area (local or state representatives) conducts a damage assessment to estimate the amount of support required from the federal government. That assessment goes through the state government to the Federal Emergency Management Agency (FEMA) to the Department of Homeland Security (DHS) to the US President, who ultimately decides whether the affected region should receive an emergency declaration, a major disaster declaration, or neither [@mccarthy_femas_2009]. FEMA then deploys resources from across the federal government to support the disaster region, the volume and type of which depend upon the declaration type. However, the federal government does not necessarily incur all of the cost of these resources. Under most circumstances, this pay structure more accurately reflects a reimbursement model; that is, the states or territories submit requests to the FEMA for reimbursement after the expenses have occurred. However, there is then a reconciliation process by which the total funds a state or territory receives is often different than the original request. The funding process was further refined as part of the Sandy Recovery Improvement Act of 2013 as well as The Disaster Recovery Reform Act of 2019 [@fema_guide_2021]. These laws require FEMA to publicly provide the information regarding federal government resources paid for by state and federal government funds for each disaster declaration. As a result, the OpenFEMA Data Sets website now exists and provides information summarizing disaster declarations and mission assignments.

While these acts have improved the accessible information, little has been done to draw conclusions from these data for predicting the financial impacts of disasters. Due to the federal government legislative process, these datasets have been publicly available for less than two years, during which the preponderance of the emergency management community was dedicated to managing the COVID-19 pandemic. To date, there is one published paper using predictive modeling to examine disaster declarations; however, it is only published in Spanish, and the authors do not have a reliable means of translation [@Perez]. Based on an approximate translation, the authors used a logistic regression to predict whether or not a presidential declaration would be awarded in the state of Maryland, utilizing Maryland data 2003 - 2017 [@Perez]. Given this newly available data from the federal government, there are numerous potential predictive models to be considered.

@@

@@

## Model Development
Four models were considered for the prediction analysis, and each were fit for both outcomes of interest (e.g. requested and obligated funds). The data were split such that 66.7% were used for training the models and the remaining 33.3% were used for testing the final chosen model. Spatial cross-validation was used for all models, using a five-fold resampling structure. The tidymodels framework utilized in the analysis does not support a repreated spatial cross-validation resampling structure, as is standard in typical cross-validation.

## Model Definition
There are numerous machine learning models that could be applied to the research question, but this analysis focuses primarily on four different types of models. The tree-based models were chosen as the primary users of the prediction algorithms are non-statisticians, and such models are among the easiest machine learning models to interpret and understand [@james_introduction_2021]. However, to robustly consider the machine learning methodology, two regularization-based methods were also included in the analysis.

* Random Forest (RF): A slightly more sophisticated model, a RF model aims to reduce the variance by building a tree for each re-sampling of the existing data. However, instead of considering all possible predictors at each decision tree split, the split occurs based on the best predictors of a random sample of all predictors. This method allows for de-correlation in the trees by avoiding the default nature of decision trees that aim to include as many predictors as possible. Each individual tree is then averaged to find the final model. RF models are also more difficult to interpret than standard decision tree models [@breiman_random_2001].

* Elastic Net (EN):

* LASSO:
